Using Model Checking to Analyze the System Behavior of the LHC Production Grid

### 2013-02-15 FGCS_rebuttal.txt

We value the effort invested by the reviewers in assessing our paper.
A list of changes for the raised points:
-------------
* [Introduction clearly describes one of the main interests of the paper: a methodology for building models (in the model-checking sense) for a legacy system. On the other hands, it fails to introduce the second contribution of the paper, which is a methodology to actually performing the check...]

> Introduction has been extended, paragraph:
[Once the model is in place, the toolset can automatically check for general system properties, such as deadlocks. We show how to formulate and check application-specific properties... 
This idea is simple but flexible and generic enough to describe many temporal properties of concurrent systems.]
-------------
* [However, for a reader not familiar with the history of the EGEE/EGI/WLCG grid, some information is missing. Mainly, Dirac appears as a standalone system, and the underlying software and hardware infrastructure is not described.
- The paper should explain the relation of Dirac with the gLite middleware, and provide a reference to  glite (eg E. Laure, S.M. Fisher, A. Frohner, C. Grandi, P. Kunszt, et al. Programming the Grid with gLite. Computational Methods in Science and Technology, 12(1):33-45, 2006.).]

@@

> This explanation has been added in Section 2.2, paragraph:
[Originally, the DIRAC WMS submitted pilot jobs using the native gLite middleware tools, but this method was shown to be suboptimal. The gLite resource brokers... making sites compete with each other for jobs and have a faster job turnaround.]
> gLite reference also added
-------------
* [The paper should give figures showing the size and scale of distribution of the hardware infrastructure (number of sites, of CPUs or cores, storage capacity etc.) accessible to DIRAC.]
> Paragraph added:
[It provides a concurrent use of over 60K CPUs and 10M file replicas distributed over tens of grid sites, ....About 10PB of data is stored and visible in the LHCb file catalogs, with daily data transfers between storage units reaching 3Gbps.] 
> Figures 1 and 2 added for illustration.
-------------
* [Considering the SMS, a state  machine analogous to the one for WMS would improve the readability of section 2.3.]
> Figure 7 added.
-------------
* [ Other approaches of model checking (eg deductive verification) could have been discussed. Also, the choice of model checking versus trace mining (a very active research area) should be justified, as both Dirac and gLite services generate copious logs that were not exploited for diagnosis. Overall, a Related Work section is necessary.]
> Related work was addressed by all reviewers, so we dedicated a separate Section 5 for this.

* [Section 4 is also very pedagogical: it follows the real-world checking process, including preliminary visualization. However, figures 7 should be provided with a better resolution and better explained ... Figure 8 should probably printed in color, or entirely removed. If kept, the "suspicious transition" should be pointed in the figure.]
> Figure 7 (now Figure 10) is now with a better resolution (background circles are removed. They simply indicate which portion of the entire state space belongs to this self-transitions. We found it un-necessary to bother the reader with this detail). Figure 8 (now Figure 11): suspicious transition is clearly marked now.
-------------
* [Section 4.3 demonstrates the efficiency of the approach, on the smallest of the two state machine. On the other hand, the approach fails on the WMS, due to combinatorial explosion.  The strategy to cope with this issue is presented as a "trick", and should be better related to the relevant concepts in the literature, in particular to testing.]
> We briefly relate this in Section 5. Related Work [This component resembles our monitoring process automaton used to cope with state space explosion during model checking of DIRAC’s WMS system]
-------------
* [The authors motivate the problem of lost efficiency of resource usage and manpower. Are there any statistics showing actual loss of efficiency? How often the DIRAC system get into inconsistent state?]
> Added in Introduction: [Depending on the frequency of certain heavier data (re)processing campaigns throughout the year, the occurrence of such inconsistencies can be up to several reported cases per month..., compared to a temporary fix, like manipulating database records or restarting processes manually]
-------------
* [The application of model checking on DIRAC system and finding race conditions is useful. However, the extendability of this approach for other systems is not clear to me. Can these approaches be applied on other systems? What changes need to be done?]
> We attempt to tackle this in Section 6. Conclusions. A few more lines of explanation are added: [It is observed that many large concurrent/distributed systems are essentially parametrized, meaning that they contain components or process descriptions which are behaviorally similar...and we believe that the combination of code slicing and abstraction rules of thumb presented in Section 3 can be helpful in this direction.]
-------------
* Reviewer #4's comments again raise the issues of applicability of the approach to other systems, and more in depth discussion of related work. (covered above)
-------------
* [Some discussion on the complexity and/or time requirements of using the model would be interesting to understand the usefulness of the model]
> This is difficult to generalize for any model, but we provide statistics on the state space generation and model checking times in section 4.Analysis and Issues.

Best regards

### 2013-02-15 ElsevierFGCS_singleColumn.tex
..

\begin{figure}[tp!]
\includegraphics[width=0.7\textwidth]{./graphics/Figure1_1.png}
\centering
\caption{LHCb job troughput by site}
\label{fig:Job-throughput}
\end{figure}
\begin{figure}[tp!]
\includegraphics[width=0.7\textwidth]{./graphics/Figure1_2.png}
\centering
\caption{LHCb data transfers throughput}
\label{fig:Data-transfers}
\end{figure}
The development of DIRAC started in 2002 as a system for production of simulation
data that would serve to verify physics theory, aspects of the LHCb detector design, as
well as to optimize physics algorithms. It gradually evolved into a complete community
grid solution for data and job management, based on a general-purpose framework
that can be reused by other communities besides LHCb. Today, it covers all major
LHCb tasks starting with the raw data transfer from the experiment's detector to
the grid storage, several steps of data processing, up to the final user
analysis. 
It provides a concurrent use of over 60K CPUs and 10M file replicas distributed
over
tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The
community of users has grown to several hundreds, loading the 
LHC grid resources with up to 30K simultaneously running jobs
(Fig.~\ref{fig:Job-throughput}) and over 100K jobs executed per day,
during peak processing periods.
About 10PB of data is stored and visible in the LHCb file catalogs, with daily
data transfers between storage units reaching 3Gbps
(Fig.~\ref{fig:Data-transfers}).

@@

\textit{Agents} are active components that fulfill a limited number of specific system
functions. They run in different environments, depending on their mission.
Some are deployed close to the corresponding services, while others run on the
grid worker nodes. Examples of the latter are the so-called Pilot Agents, part
of the Workload Management System explained in the following section. All DIRAC
agents repeat the same logic in each iteration cycle: they monitor the service states,
 and respond by initiating actions (like job
submission or data transfer) which update the states of various system
entities.
\begin{figure}[tp!]
\includegraphics[width=0.5\textwidth] {./graphics/Figure1.png}
\centering
\caption{DIRAC Architecture overview}
\label{fig:DIRAC-Arch}
\end{figure}
\textit{Resources} are software abstractions of the underlying heterogeneous grid
computing and storage entities allocated to LHCb, providing a uniform interface
for access. The physical resources are controlled by the site managers and made
available through middleware services such as gLite \cite{Laure06programmingthe}.

@@

The driving force of DIRAC is the Workload Management System (WMS). Taking into
account the heterogeneous, dynamic, and high-latency nature of the distributed computing environment, a
\textit{Pilot Job paradigm}, illustrated in Fig.~\ref{fig:DIRAC-WMS}, 
\begin{figure}[bp]
\includegraphics[width=0.7\textwidth]{./graphics/Figure2.png}
\centering
\caption{DIRAC Workload Management System \cite{DIRAC_pilot_WMS}}
\label{fig:DIRAC-WMS}
\end{figure}
was chosen as an efficient way to
implement a pull scheduling mechanism.  Pilot jobs are simply resource
reservation processes without an actual payload defined a priori. They are
submitted to the grid worker nodes with the aim of checking the sanity of the
operational environment just before pulling and executing the real payload. This
hides the fragility of the underlying resources and increases the job success
rate, from the perspective of end users. Centralized \textit{Task Queues} contain all the
pending jobs, organized in groups according to their requirements. This enables
efficient application of job priorities, enforcing fair-share policies and
quotas, which are decided within the LHCb community, without the need of any
effort from the grid sites.  Originally, the DIRAC WMS submitted pilot jobs using 
the native gLite middleware tools, but this method was shown to be suboptimal. The gLite resource brokers
are centralized components, and as such they do not have the capacity and flexibility to cope with the load of 
a community such as LHCb. Using multiple resource brokers, rather than a single one, was not the solution:
all brokers use the same site state information and would always submit jobs to the most suitable one globally, causing overloads to
certain sites while underloading others very often. In addition, the gLite information system was not reactive enough to propagate
changes in site status information to the brokers. With the wide deployment of the CREAM CE service \cite{Aiftimiei2010654}, DIRAC's WMS switched
to submitting pilot jobs directly to sites. By interrogating the Computing Element status directly, the system reactiveness
improved dramatically, making sites compete with each other for jobs and have a faster job turnaround.

The basic flowchart describing the evolution of a job's states is depicted in
Fig.~\ref{fig:DIRAC-job-state-machine}. After submission through the \textit{Job Manager} (``Received''), the complete
job description is placed in the DIRAC job repository (the Job DB). Before jobs
become eligible for execution, a chain of optimizer agents checks and prioritizes
them in queues, utilizing the parameters information from the Job DB.
\begin{figure}[tp]
\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure3.png}
\centering
\caption{Job state machine \cite{ProductionShifterGuide}}
\label{fig:DIRAC-job-state-machine}
\end{figure} 
If the requested data resides on tape storage, the \textit{Job
Scheduling Agent} will pass the control to a specialized \textit{Stager} service (part of
the SMS explained in the next section), before placing the job in a Task Queue
(``Waiting''). Based on the complete list of pending payloads, a specialized \textit{Task
Queue Director} agent submits pilots to the computing resources via the available job
submission middleware (e.g., gLite WMS). After
a \textit{Matcher} service pulls the most suitable payload for a pilot (``Matched''), a \textit{Job
Wrapper} object is created on the worker node, responsible for retrieving the
input sandbox, performing software availability checks, executing the actual
payload on the worker node (``Running''), and finally uploading any output data
necessary (``Done'' or ``Failed''). The wrapper can catch any failure exit state
reported by the running physics applications. At the same time, a \textit{Watchdog}
process is instantiated to monitor the behavior of the Job Wrapper and send
heartbeat signals to the monitoring service. It can also take actions in case
resources are soon to be exhausted, the payload stalls, or a management command
for killing the payload is received.

@@

\begin{figure}
\centering
\begin{minipage}{.5\textwidth}
  \centering
  \includegraphics[width=0.98\linewidth]{./graphics/Figure4.png}
  \caption{DIRAC Storage Management System}
  \label{fig:DIRAC-SMS}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
  \includegraphics[width=0.95\linewidth]{./graphics/SMS_CacheReplicasSM.png}
  \caption{CacheReplicas state machine}
  \label{fig:CacheReplicas-SM}
\end{minipage}
\end{figure}

@@

Abstraction aims at reducing
the program's state space in order to overcome the resource limitations \cite{Pelánek08fightingstate} encountered during model-checking. 
Furthermore, specification languages describe \textit{what} is being done, 
abstracting away from the details of \textit{how} things are done. 
Then, the ultimate question is: \textit{how do we establish correspondence between model and code?}

@@

\subsubsection{Data Abstractions}
The CacheReplicas entity contains more information besides the status of a file.
Every database entry has a unique identifier, descriptive data such as 
the storage where it resides, its full path, checksum, timestamps, etc., as can be seen
from the table description in Fig.~\ref{fig:CacheReplicasTable}.
\begin{figure}[bp!]
\begin{center}
 \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure5.png}
\end{center}
\caption{CacheReplicas table description}
\label{fig:CacheReplicasTable}
\end{figure}
Model checking can only be performed on closed models, where the domains
of all variables are finite. Since we are only interested in state transitions in terms of 
changes to the \textit{Status} field,
we can collapse most of this descriptive data, and represent this entity as a 
\textit{user-defined sort} (type) in mCRL2:
\begin{displaymath}
\begin{aligned}
sort\ CacheReplicas = \textbf{struct}\ Start\ | 
			  New\ |  \\
		    Waiting\ | 
          StageSubmitted\ | \\
		      Staged\ | 
		      Failed\ | 
		  Deleted\ ;
\end{aligned}
\end{displaymath}
This defines an enumerated data type with all possible states.
The Tasks entity is modeled in the same manner. Lists of these sorts can be easily modeled 
in mCRL2 as \begin{math}List(Tasks) \end{math} and 
\begin{math}List(CacheReplicas) \end{math}.
To define the many-to-many relationship between Tasks and CacheReplicas,
we join these data elements in a tuple:
%\begin{lstlisting}[tabsize=10,basicstyle=\ttfamily\fontsize{8}{7}\selectfont]
\begin{displaymath}sort \ Tuple = \textbf{struct}\ p(t:Nat,r:Nat,link:Bool)
\end{displaymath}
%\end{lstlisting}
The first two elements are the list positions of the Tasks and CacheReplicas
entries, while the last one indicates whether a relation between
them exists at a given moment of the system execution.

@@

  \subfloat[Logging info of a DIRAC job]{\label{fig:Job-Logging-Info}\includegraphics[width=0.4566\linewidth]{./graphics/Figure6a.png}}    

@@

  \subfloat[XSim simulator trace]{\label{fig:simulator}\includegraphics[width =0.5\linewidth]{./graphics/Figure6b.png}}

@@

\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure7_new.png}

@@

\includegraphics[width=0.57\linewidth,keepaspectratio=true]{./graphics/Figure8.png}

@@

\begin{figure}[tp]
\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure9.png}
%was PropertyViolation1.png
\centering
\caption{Violation of requirements 1 (top) and 2 (bottom)}
\label{fig:PropertyViolation}
\end{figure}
Model checking was carried out on a 64bit Intel Core\texttrademark2 Duo (1.6GHz) machine with 2GB RAM. 
Both requirements 1 and 2 were found to be violated, as can be seen from the traces (Fig.~\ref{fig:PropertyViolation}).
The race conditions manifest themselves in a few subtle interactions between the SMS storage and the agents.
In both cases, at step 4 the \textit{Stage Request Agent} issues prestage requests and 
as a result the CacheReplicas entries (second element of the State column) are updated to ``StageSubmitted''. Between the moment
that this agent has selected the corresponding Task to update to the same state,
to the point where the update is done (last step in both traces),
other agents may have monitored these replica records and updated
them to a new state, along with the associated Task. In practice, the manifestation
of such race conditions depends on the speed of the agents propagating the
state changes between the selection and update done by the \textit{Stage Request Agent}.
Such cases are nevertheless encountered in reality, when this agent
has large lists of records to process. This results in a deadlock situation, as 
the Task will have no further state updates made by other agents, since
from their perspective the state propagation is finished.

@@

The automaton of the concrete monitor instance which accepts requirement 3 trace is shown in Fig.~\ref{fig:monitorProcess}. 
\begin{figure}[tp]
\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure10.png}
\centering
\caption{Monitor automaton for requirement 3}
\label{fig:monitorProcess}
\end{figure}%123478
Whenever the observed system changes the state of a job to ``Killed'', the automaton also changes
its state to L1 by enforcing synchronous communication between the respective actions.
The silent step (${\tau}$) can be executed as a result of any non-relevant system state change
from the perspective of the automaton. As soon as the system changes the state of the job further
to ``Running'', the automaton makes a move in a similar fashion, after which point the action 
\textit{error} is executed, and can be detected by the LTSmin tool. 
\begin{figure}[bp]
\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure11.png}
\centering
\caption{"Zombie" job starts running after being killed}
\label{fig:zombieJob}
\end{figure}%123478
If such a full cycle is accepted by the automaton, the temporal property is indeed violated.
Using this technique, a counterexample (Fig.~\ref{fig:zombieJob}) showed that, when 
staging was involved in the workflow of a job, the callback from
the SMS was not properly handled.
It awakened the job immediately to a ``Checking'' (and eventually ``Running'') state
even if it had been manually ``Killed'' 
(using DIRAC's command line API) by production managers meanwhile. Such zombie jobs
were discovered in the system on several occasions,
and with the model at hand it was easy to replay the behavior and localize the
problem. The implementation was fixed to properly guard the callback against this particular case.
The changes were also reflected back to the mCRL2 model ~\cite{svn_mcrl2}. Fig.~\ref{fig:zombieJobFix}
shows that the stager callback no longer affects the job state if it has been changed from its
original ``Staging'' one in the meantime. 
The state space of the fixed model amounts to 
157048664 states, the reduction being a result of the blocking of the unwanted behavior.
However, we must note that model checking still discovered traces where the property 
is violated, even without the stager being involved.
The current architecture does not permit an easy workaround for this problem.
Protecting shared memory access using traditional locking techniques does not scale well.
There is an ongoing effort in redesigning DIRAC's architecture, to accomodate
a task dispatcher that takes care of persisting the state of the tasks to the storage 
backend and distributing them among so called ``executors''.This should 
replace the old agent components' polling architecture, and reduce the load
they place on the storage, in addition to eliminating race conditions.

%TODO: fix the model, show a fixed trace
\begin{figure}[tp]
\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure12.png}
\centering
\caption{ Race condition involving the stager callback no longer occurs}
\label{fig:zombieJobFix}
\end{figure}%123478

>>>>

\begin{figure}[bp!]
\includegraphics[width=0.7\textwidth]{./graphics/Figure1.png}
\centering
\caption{LHCb job troughput by site}
\label{fig:Job-throughput}
\end{figure}
\begin{figure}[tp!]
\includegraphics[width=0.7\textwidth]{./graphics/Figure2.png}
\centering
\caption{LHCb data transfers throughput}
\label{fig:Data-transfers}
\end{figure}
The development of DIRAC started in 2002 as a system for production of simulation
data that would serve to verify physics theory, aspects of the LHCb detector design, as
well as to optimize physics algorithms. It gradually evolved into a complete community
grid solution for data and job management, based on a general-purpose framework
that can be reused by other communities besides LHCb. Today, it covers all major
LHCb tasks starting with the raw data transfer from the experiment's detector to
the grid storage, several steps of data processing, up to the final user
analysis. 
It provides a concurrent use of over 60K CPUs and 10M file replicas distributed
over
tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The
community of users has grown to several hundreds, loading the 
LHC grid resources with up to 30K simultaneously running jobs
(Fig.~\ref{fig:Job-throughput}) and over 100K jobs executed per day,
during peak processing periods.
About 10PB of data is stored and visible in the LHCb file catalogs, with daily
data transfers between storage units reaching 3Gbps
(Fig.~\ref{fig:Data-transfers}).

@@

\textit{Agents} are active components that fulfill a limited number of specific system
functions. They run in different environments, depending on their mission.
Some are deployed close to the corresponding services, while others run on the
grid worker nodes. Examples of the latter are the so-called Pilot Agents, part
of the Workload Management System explained in the following section. All DIRAC
agents repeat the same logic in each iteration cycle: they monitor the service states,
 and respond by initiating actions (like job
submission or data transfer) which update the states of various system
entities.
\begin{figure}[bp!]
\includegraphics[width=0.5\textwidth] {./graphics/Figure3.png}
\centering
\caption{DIRAC Architecture overview}
\label{fig:DIRAC-Arch}
\end{figure}
\textit{Resources} are software abstractions of the underlying heterogeneous grid
computing and storage entities allocated to LHCb, providing a uniform interface
for access. The physical resources are controlled by the site managers and made
available through middleware services such as gLite \cite{Laure06programmingthe}.

@@

The driving force of DIRAC is the Workload Management System (WMS). Taking into
account the heterogeneous, dynamic, and high-latency nature of the distributed computing environment, a
\textit{Pilot Job paradigm}, illustrated in Fig.~\ref{fig:DIRAC-WMS}, 
\begin{figure}[bp]
\includegraphics[width=0.7\textwidth]{./graphics/Figure4.png}
\centering
\caption{DIRAC Workload Management System \cite{DIRAC_pilot_WMS}}
\label{fig:DIRAC-WMS}
\end{figure}
was chosen as an efficient way to
implement a pull scheduling mechanism.  Pilot jobs are simply resource
reservation processes without an actual payload defined a priori. They are
submitted to the grid worker nodes with the aim of checking the sanity of the
operational environment just before pulling and executing the real payload. This
hides the fragility of the underlying resources and increases the job success
rate, from the perspective of end users. Centralized \textit{Task Queues} contain all the
pending jobs, organized in groups according to their requirements. This enables
efficient application of job priorities, enforcing fair-share policies and
quotas, which are decided within the LHCb community, without the need of any
effort from the grid sites.  Originally, the DIRAC WMS submitted pilot jobs using 
the native gLite middleware tools, but this method was shown to be suboptimal. The gLite resource brokers
are centralized components, and as such they do not have the capacity and flexibility to cope with the load of 
a community such as LHCb. Using multiple resource brokers, rather than a single one, was not the solution:
all brokers use the same site state information and would always submit jobs to the most suitable one globally, causing overloads to
certain sites while underloading others very often. In addition, the gLite information system was not reactive enough to propagate
changes in site status information to the brokers. With the wide deployment of the CREAM CE service \cite{Aiftimiei2010654}, DIRAC's WMS switched
to submitting pilot jobs directly to sites. By interrogating the Computing Element status directly, the system reactiveness
improved dramatically, making sites compete with each other for jobs and have a faster job turnaround.

The basic flowchart describing the evolution of a job's states is depicted in
Fig.~\ref{fig:DIRAC-job-state-machine}. After submission through the \textit{Job Manager} (``Received''), the complete
job description is placed in the DIRAC job repository (the Job DB). Before jobs
become eligible for execution, a chain of optimizer agents checks and prioritizes
them in queues, utilizing the parameters information from the Job DB.
\begin{figure}[tp]
\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure5.png}
\centering
\caption{Job state machine \cite{ProductionShifterGuide}}
\label{fig:DIRAC-job-state-machine}
\end{figure} 
If the requested data resides on tape storage, the \textit{Job
Scheduling Agent} will pass the control to a specialized \textit{Stager} service (part of
the SMS explained in the next section), before placing the job in a Task Queue
(``Waiting''). Based on the complete list of pending payloads, a specialized \textit{Task
Queue Director} agent submits pilots to the computing resources via the available job
submission middleware (e.g., gLite WMS). After
a \textit{Matcher} service pulls the most suitable payload for a pilot (``Matched''), a \textit{Job
Wrapper} object is created on the worker node, responsible for retrieving the
input sandbox, performing software availability checks, executing the actual
payload on the worker node (``Running''), and finally uploading any output data
necessary (``Done'' or ``Failed''). The wrapper can catch any failure exit state
reported by the running physics applications. At the same time, a \textit{Watchdog}
process is instantiated to monitor the behavior of the Job Wrapper and send
heartbeat signals to the monitoring service. It can also take actions in case
resources are soon to be exhausted, the payload stalls, or a management command
for killing the payload is received.

@@

\begin{figure}
\centering
\begin{minipage}{.5\textwidth}
  \centering
  \includegraphics[width=0.98\linewidth]{./graphics/Figure6.png}
  \caption{DIRAC Storage Management System}
  \label{fig:DIRAC-SMS}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
  \includegraphics[width=0.95\linewidth]{./graphics/Figure7.png}
  \caption{CacheReplicas state machine}
  \label{fig:CacheReplicas-SM}
\end{minipage}
\end{figure}

@@

Abstraction aims at reducing
the program's state space in order to overcome the resource limitations \cite{Pelanek08fightingstate} encountered during model-checking. 
Furthermore, specification languages describe \textit{what} is being done, 
abstracting away from the details of \textit{how} things are done. 
Then, the ultimate question is: \textit{how do we establish correspondence between model and code?}

@@

\subsubsection{Data Abstractions}
The CacheReplicas entity contains more information besides the status of a file.
Every database entry has a unique identifier, descriptive data such as 
the storage where it resides, its full path, checksum, timestamps, etc., as can be seen
from the table description in Fig.~\ref{fig:CacheReplicasTable}.
\begin{figure}[bp!]
\begin{center}
 \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
\end{center}
\caption{CacheReplicas table description}
\label{fig:CacheReplicasTable}
\end{figure}
Model checking can only be performed on closed models, where the domains
of all variables are finite. Since we are only interested in state transitions in terms of 
changes to the \textit{Status} field,
we can collapse most of this descriptive data, and represent this entity as a 
\textit{user-defined sort} (type) in mCRL2:
\begin{displaymath}
\begin{aligned}
sort\ CacheReplicas = \textbf{struct}\ Start\ | 
			  New\ |  \\
		    Waiting\ | 
          StageSubmitted\ | \\
		      Staged\ | 
		      Failed\ | 
		  Deleted\ ;
\end{aligned}
\end{displaymath}
This defines an enumerated data type with all possible states.
The Tasks entity is modeled in the same manner. Lists of these sorts can be easily modeled 
in mCRL2 as \begin{math}List(Tasks) \end{math} and 
\begin{math}List(CacheReplicas) \end{math}.
To define the many-to-many relationship between Tasks and CacheReplicas,
we join these data elements in a tuple:
%\begin{lstlisting}[tabsize=10,basicstyle=\ttfamily\fontsize{8}{7}\selectfont]
\begin{displaymath}sort \ Tuple = \textbf{struct}\ p(t:Nat,r:Nat,link:Bool)
\end{displaymath}
%\end{lstlisting}
The first two elements are the list positions of the Tasks and CacheReplicas
entries, while the last one indicates whether a relation between
them exists at a given moment of the system execution.

@@

  \subfloat[Logging info of a DIRAC job]{\label{fig:Job-Logging-Info}\includegraphics[width=0.4566\linewidth]{./graphics/Figure9a.png}}    

@@

  \subfloat[XSim simulator trace]{\label{fig:simulator}\includegraphics[width =0.5\linewidth]{./graphics/Figure9b.png}}

@@

\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure10.png}

@@

\includegraphics[width=0.57\linewidth,keepaspectratio=true]{./graphics/Figure11.png}

@@

\begin{figure}[tp]
\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure12.png}
%was PropertyViolation1.png
\centering
\caption{Violation of requirements 1 (top) and 2 (bottom)}
\label{fig:PropertyViolation}
\end{figure}
Model checking was carried out on a 64bit Intel Core\texttrademark2 Duo (1.6GHz) machine with 2GB RAM. 
Both requirements 1 and 2 were found to be violated, as can be seen from the traces (Fig.~\ref{fig:PropertyViolation}).
The race conditions manifest themselves in a few subtle interactions between the SMS storage and the agents.
In both cases, at step 4 the \textit{Stage Request Agent} issues prestage requests and 
as a result the CacheReplicas entries (second element of the State column) are updated to ``StageSubmitted''. Between the moment
that this agent has selected the corresponding Task to update to the same state,
to the point where the update is done (last step in both traces),
other agents may have monitored these replica records and updated
them to a new state, along with the associated Task. In practice, the manifestation
of such race conditions depends on the speed of the agents propagating the
state changes between the selection and update done by the \textit{Stage Request Agent}.
Such cases are nevertheless encountered in reality, when this agent
has large lists of records to process. This results in a deadlock situation, as 
the Task will have no further state updates made by other agents, since
from their perspective the state propagation is finished.

@@

The automaton of the concrete monitor instance which accepts requirement 3 trace is shown in Fig.~\ref{fig:monitorProcess}. 
\begin{figure}[tp]
\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure13.png}
\centering
\caption{Monitor automaton for requirement 3}
\label{fig:monitorProcess}
\end{figure}%123478
Whenever the observed system changes the state of a job to ``Killed'', the automaton also changes
its state to L1 by enforcing synchronous communication between the respective actions.
The silent step (${\tau}$) can be executed as a result of any non-relevant system state change
from the perspective of the automaton. As soon as the system changes the state of the job further
to ``Running'', the automaton makes a move in a similar fashion, after which point the action 
\textit{error} is executed, and can be detected by the LTSmin tool. 
\begin{figure}[tp]
\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure14.png}
\centering
\caption{"Zombie" job starts running after being killed}
\label{fig:zombieJob}
\end{figure}%123478
If such a full cycle is accepted by the automaton, the temporal property is indeed violated.
Using this technique, a counterexample (Fig.~\ref{fig:zombieJob}) showed that, when 
staging was involved in the workflow of a job, the callback from
the SMS was not properly handled.
It awakened the job immediately to a ``Checking'' (and eventually ``Running'') state
even if it had been manually ``Killed'' 
(using DIRAC's command line API) by production managers meanwhile. Such zombie jobs
were discovered in the system on several occasions,
and with the model at hand it was easy to replay the behavior and localize the
problem. The implementation was fixed to properly guard the callback against this particular case.
The changes were also reflected back to the mCRL2 model ~\cite{svn_mcrl2}. Fig.~\ref{fig:zombieJobFix}
shows that the stager callback no longer affects the job state if it has been changed from its
original ``Staging'' one in the meantime. 
The state space of the fixed model amounts to 
157048664 states, the reduction being a result of the blocking of the unwanted behavior.
However, we must note that model checking still discovered traces where the property 
is violated, even without the stager being involved.
The current architecture does not permit an easy workaround for this problem.
Protecting shared memory access using traditional locking techniques does not scale well.
There is an ongoing effort in redesigning DIRAC's architecture, to accomodate
a task dispatcher that takes care of persisting the state of the tasks to the storage 
backend and distributing them among so called ``executors''.This should 
replace the old agent components' polling architecture, and reduce the load
they place on the storage, in addition to eliminating race conditions.

%TODO: fix the model, show a fixed trace
\begin{figure}[tp]
\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure15.png}
\centering
\caption{ Race condition involving the stager callback no longer occurs}
\label{fig:zombieJobFix}
\end{figure}%123478
